Fractional Decomposition Tree Algorithm: A tool for studying the integrality gap of Integer Programs

### 2020-08-09 FDT2020/main-section.tex
typo in comment
Removing the big comment.  Fixes related to my earlier comments/confusion.

Consider an IP instance $I=(A,b)$. Define sets $S(I)$ and $P(I)$ as in (\ref{S}) and (\ref{P}), respectively. Assume $S(I)\subseteq \{0,1\}^n$ and $P(I)\subseteq [0,1]^n$. For simplicity in the notation we assume an instance $I$ and denote $P(I),S(I),$ and $g(I)$ by $P$, $S$, and $g$ for this section and the next section. Also, for both sections let $x^*$ be the optimal solution to the LP formulation and assume $t=|\spp(x^*)|$. Without loss of generality we can assume $x^*_i = 0$ for $i=t+1,\ldots,n$.
%\cindy{TODO: Is this something that should be clear to the reader with what they know at this point? Does it need an explanation? Does it need an explanation at the end of this section or the next?}\arash{this is just a relabeling of the indices.} \cindy{Sorry.  I didn't understand that you meant the solution has that property. Right? The vector $x$ isn't defined in this paragraph and it's not in the next lemma either. If $x$ is the output of DomToIP, then we should say that.  ``Without loss of generality the solution $x$ returned by DomToIP has ...''}\arash{I see your point, please see if the change made it better} \cindy{Yes, thanks.  We can delete this conversation after you see this response.}

In this section we prove Theorem \ref{domtoIP}. In fact, we prove a stronger result. 
\begin{lemma}\label{domlemma}
	Given $\tilde{x}\in \dom(P)$ and $\tilde{x}\in \{0,1\}^n$, there is an algorithm (the DomToIP algorithm) that finds $\bar{x}\in S$ in polynomial time, such that $\bar{x}\leq \tilde{x}$.\end{lemma}
Lemma \ref{domlemma} implies Theorem \ref{domtoIP}, since it is easy to obtain an integer point in $\dom(P)$: numerically rounding up $x^*$ (or any fractional point in $P$) gives us a point in $\dom(P)$.
\cindy{Does this Lemma need the caveat that the instance has a finite integrality gap?}
Hence, we can assume in the proof below that $\tilde{x}_i= 0$ for $i=t+1,\ldots,n$. 
%\arash{I also made a change based on the change above}\cindy{Thanks!}
\subsection{Proof of Lemma \ref{domlemma}: The DomToIP Algorithm}

We start by introducing an algorithm that ``fixes" the variables iteratively, starting from the first coordinate and ending at the $t$-th coordinate. Suppose we run the algorithm for $\ell\in \{0,\ldots,t-1\}$ iterations and in each iteration we find $x^{(\ell)}\in \dom(P)$  such that $x^{(\ell)}_i\in \{0,1\}$ for $i=1,\ldots,\ell$. We can set $x^{(0)}=\tilde{x}$. Now consider the following linear program. The variables of this linear program are the $z\in \mathbb{R}^n$ variables.
\begin{align}
\HelperLP(x^{(\ell)})\quad\quad& \min\quad \;z_{\ell+1}\\
&\;\text{s.t.} \quad \;\;Az\geq b \\
&\;{\color{white}{\text{s.t.}} }\quad \;\; \; z_j = x^{(\ell)}_j \quad \; j =1,\ldots, \ell\\
&\;{\color{white}{\text{s.t.}} }\quad \; \;\; z_j \leq x^{(\ell)}_j \quad \; j = \ell+1,\ldots,n\\
&\;{\color{white}{\text{s.t.}} }\quad \; \;\; z\;\geq 0
\end{align}

@@

We prove that indeed $x^{(t)}\in S$. First, we need to show that in any iteration $\ell=  0,\ldots,t-1$ of DomToIP that $\HelperLP(x^{(\ell)})$ is feasible. We show something stronger. For $\ell=0,\ldots,t-1$ let
\begin{align*}
\LP^{(\ell)}&= \{z\in P\; : \; z\leq x^{(\ell)} \mbox{ and } z_j=x_j^{(\ell)} \mbox{ for } j\in [\ell]\}, \text{ and}\\
\IP^{(\ell)}&= \{z\in \LP^{(\ell)}\; : \; z\in \{0,1\}^n\}.
\end{align*}
If $\LP^{(\ell)}$ is a non-empty set then $\HelperLP(x^{(\ell)})$ is feasible. We show by induction on $\ell$ that $\LP^{(\ell)}$ and $\IP^{(\ell)}$ are not empty sets for $\ell=0,\ldots,t-1$. First notice that $\LP^{(0)}$ is clearly feasible since by definition $x^{(0)}\in \dom(P)$, meaning there exists $z\in P$ such that $z\leq x^{(0)}$. By Theorem \ref{CV2}, there exists $\tilde{z}^i\in S$ and $\theta_i\geq 0$ for $i\in [k]$ such that $\sum_{i=1}^{k} \theta_i = 1$ and $\sum_{i=1}^{k}\theta_i \tilde{z}^i \leq gz$. Hence, $\sum_{i=1}^{k}\theta_i \tilde{z}^i \leq gz\leq gx^{(0)}$. So if $x^{(0)}_j=0$, then $ \sum_{i=1}^{k}\theta_i \tilde{z}_j^i =0$, which implies that $\tilde{z}^i_j=0$ for all $i\in [k]$ and $j\in [n]$ where $x^{(0)}_j=0$. Hence, $\tilde{z}^i\leq x^{(0)}$ for $i\in [k]$. Therefore $\tilde{z}^i\in \IP^{(0)}$ for $i\in [k]$, which implies $\IP^{(0)}\neq \emptyset$.

@@

We consider two cases. In the first case, we have $z^*_{\ell+1}=0$. In this case we have $x^{(\ell+1)}_{\ell+1}=0$. Since $z^*\leq x^{(\ell+1)}$, we have $z^*\in \LP^{(\ell+1)}$. Also, $z^*\in P$. By Theorem \ref{CV2} there exists $\tilde{z}^i\in S$ and $\theta_i\geq 0$ for $i\in [k]$ such that $\sum_{i=1}^{k} \theta_i = 1$ and  $\sum_{i=1}^{k}\theta_i \tilde{z}^i \leq gz^*$. We have $\sum_{i=1}^{k}\theta_i \tilde{z}^i \leq gz^*\leq gx^{(\ell+1)}$.
So for $j\in [n]$ where $x^{(\ell+1)}_j=0$, we have $z^i_j=0$ for $i\in [k]$. This implies $\tilde{z}^i\leq x^{(\ell+1)}$ for $i=1,\ldots,k$. Hence, there exists $z\in S$ such that $z\leq x^{(\ell+1)}$. We claim that $z\in \IP^{(\ell+1)}$. If $z\notin \IP^{(\ell+1)}$ we must have $1\leq j \leq \ell$ such that $z_j < x^{(\ell+1)}_{j}$, and thus $z_j = 0$ and $x^{(\ell+1)}_j=1$. Without loss of generality assume $j$ is minimum number satisfying $z_j < x^{(\ell+1)}_{j}$. Consider iteration $j$ of the DomToIP algorithm. We have $z\leq x^{(\ell+1)}\leq x^{(j)}$. We also have $x^{(j)}_j=1$ which implies when we solved $\HelperLP(x^{(j-1)})$ the optimal value was strictly larger than zero. However, $z$ is a feasible solution to $\HelperLP(x^{(j-1)})$ and gives an objective value of 0. This is a contradiction, so $z\in \IP^{(\ell+1)}$.

Now for the second case, assume $z^*_{\ell+1} > 0$. We have $x^{(\ell+1)}_{\ell+1}=1$. For each point $z\in \LP^{(\ell)}$ we have $z_{\ell+1} >0$, so for each $z\in \IP^{(\ell)}$ we have $z_{\ell+1}>0$, i.e. $z_{\ell+1}=1$. This means that $z\in \IP^{(\ell+1)}$, and $\IP^{(\ell+1)} \neq \emptyset$.

>>>>

Consider an IP instance $I=(A,b)$. Define sets $S(I)$ and $P(I)$ as in (\ref{S}) and (\ref{P}), respectively. Assume $S(I)\subseteq \{0,1\}^n$ and $P(I)\subseteq [0,1]^n$. For simplicity in the notation we assume an instance $I$ and denote $P(I),S(I),$ and $g(I)$ by $P$, $S$, and $g$ for this section and the next section. Also, for both sections let $x^*$ be the optimal solution to the LP formulation and assume $t=|\spp(x^*)|$. Reorder variables as necessary to put all the nonzero variables into the first $t$ indices. So we can assume $x^*_i = 0$ for $i=t+1,\ldots,n$.
%\cindy{TODO: Is this something that should be clear to the reader with what they know at this point? Does it need an explanation? Does it need an explanation at the end of this section or the next?}\arash{this is just a relabeling of the indices.} \cindy{Sorry.  I didn't understand that you meant the solution has that property. Right? The vector $x$ isn't defined in this paragraph and it's not in the next lemma either. If $x$ is the output of DomToIP, then we should say that.  ``Without loss of generality the solution $x$ returned by DomToIP has ...''}\arash{I see your point, please see if the change made it better} \cindy{Yes, thanks.  We can delete this conversation after you see this response.}

In this section we prove Theorem \ref{domtoIP}. In fact, we prove a stronger result. 
\begin{lemma}\label{domlemma}
	Given an integral vector in the dominant of an LP relaxation ($\tilde{x}\in \dom(P)$ and $\tilde{x}\in \{0,1\}^n$) for an IP instance that has integrality gap $g < \infty$, there is an algorithm (the DomToIP algorithm) that finds $\bar{x}\in S$ in polynomial time such that $\bar{x}\leq \tilde{x}$.\end{lemma}
Lemma \ref{domlemma} implies Theorem \ref{domtoIP}, since it is easy to obtain an integer point in $\dom(P)$: numerically rounding up $x^*$ (or any fractional point in $P$) gives us a point in $\dom(P)$.
\cindy{I added the caveat that the instance has a finite integrality gap and made it a little more self-contained.  Is it OK?}
Hence, we can assume in the proof below that $\tilde{x}_i= 0$ for $i=t+1,\ldots,n$. 
%\arash{I also made a change based on the change above}\cindy{Thanks!}
\subsection{Proof of Lemma \ref{domlemma}: The DomToIP Algorithm}

We introduce an algorithm that considers each of the variables iteratively, starting from the first coordinate and ending at the $t$-th coordinate. If the initial (integral) point $\tilde{x}$ has the variable set to $1$, the algorithm tries to reduce the variable to $0$ if possible. We run the algorithm for $\ell\in \{0,\ldots,t-1\}$ iterations and in each iteration we find $x^{(\ell)}\in \dom(P)$  such that $x^{(\ell)}_i\in \{0,1\}$ for $i=1,\ldots,\ell$.
\cindy{I think the previous sentence also added to my confusion before (along with the term ``fix'' which I have replaced).  After $\ell$ iterations the first $\ell$ components are integer, but they were integer before.  Did you intend to say that we finalize a 0-1 value for the first $\ell$ coordinates?}
We can set $x^{(0)}=\tilde{x}$. Now consider the following linear program. The variables of this linear program are the $z\in \mathbb{R}^n$.
\begin{align}
\HelperLP(x^{(\ell)})\quad\quad& \min\quad \;z_{\ell+1}\\
&\;\text{s.t.} \quad \;\;Az\geq b \\
&\;{\color{white}{\text{s.t.}} }\quad \;\; \; z_j = x^{(\ell)}_j \quad \; j =1,\ldots, \ell\\
&\;{\color{white}{\text{s.t.}} }\quad \; \;\; z_j \leq x^{(\ell)}_j \quad \; j = \ell+1,\ldots,n\\
&\;{\color{white}{\text{s.t.}} }\quad \; \;\; z\;\geq 0
\end{align}

@@

We prove that indeed $x^{(t)}\in S$. First, we need to show that in any iteration $\ell=  0,\ldots,t-1$ of DomToIP that $\HelperLP(x^{(\ell)})$ is feasible. We show something stronger. For $\ell=0,\ldots,t-1$ let
\begin{align*}
\LP^{(\ell)}&= \{z\in P\; : \; z\leq x^{(\ell)} \mbox{ and } z_j=x_j^{(\ell)} \mbox{ for } j\in [\ell]\}, \text{ and}\\
\IP^{(\ell)}&= \{z\in \LP^{(\ell)}\; : \; z\in \{0,1\}^n\}.
\end{align*}
If $\LP^{(\ell)}$ is a non-empty set then $\HelperLP(x^{(\ell)})$ is feasible. We show by induction on $\ell$ that $\LP^{(\ell)}$ and $\IP^{(\ell)}$ are not empty sets for $\ell=0,\ldots,t-1$. First, $\LP^{(0)}$ is feasible since by definition $x^{(0)}\in \dom(P)$, meaning there exists $z\in P$ such that $z\leq x^{(0)}$. By Theorem \ref{CV2}, there exists $\tilde{z}^i\in S$ and $\theta_i\geq 0$ for $i\in [k]$ such that $\sum_{i=1}^{k} \theta_i = 1$ and $\sum_{i=1}^{k}\theta_i \tilde{z}^i \leq gz$. (This assumes a finite integrality gap $g$).
%\cindy{Is this the place (and others like it) where we require $g$ to be finite?}
Hence, $\sum_{i=1}^{k}\theta_i \tilde{z}^i \leq gz\leq gx^{(0)}$. So if $x^{(0)}_j=0$, then $ \sum_{i=1}^{k}\theta_i \tilde{z}_j^i =0$, which implies that $\tilde{z}^i_j=0$ for all $i\in [k]$ and $j\in [n]$ where $x^{(0)}_j=0$. Hence, $\tilde{z}^i\leq x^{(0)}$ for $i\in [k]$. Therefore $\tilde{z}^i\in \IP^{(0)}$ for $i\in [k]$, which implies $\IP^{(0)}\neq \emptyset$.

@@

We consider two cases. In the first case, we have $z^*_{\ell+1}=0$. In this case we set $x^{(\ell+1)}_{\ell+1}=0$. Since $z^*\leq x^{(\ell+1)}$, we have $z^*\in \LP^{(\ell+1)}$. Also, $z^*\in P$. By Theorem \ref{CV2} there exists $\tilde{z}^i\in S$ and $\theta_i\geq 0$ for $i\in [k]$ such that $\sum_{i=1}^{k} \theta_i = 1$ and  $\sum_{i=1}^{k}\theta_i \tilde{z}^i \leq gz^*$. We have $\sum_{i=1}^{k}\theta_i \tilde{z}^i \leq gz^*\leq gx^{(\ell+1)}$.
So for $j\in [n]$ where $x^{(\ell+1)}_j=0$, we have $z^i_j=0$ for $i\in [k]$. This implies $\tilde{z}^i\leq x^{(\ell+1)}$ for $i=1,\ldots,k$. Hence, there exists $z\in S$ such that $z\leq x^{(\ell+1)}$. We claim that $z\in \IP^{(\ell+1)}$. If $z\notin \IP^{(\ell+1)}$ we must have $1\leq j \leq \ell$ such that $z_j < x^{(\ell+1)}_{j}$, and thus $z_j = 0$ and $x^{(\ell+1)}_j=1$. Without loss of generality assume $j$ is the minimum index (between $1$ and $\ell$) satisfying $z_j < x^{(\ell+1)}_{j}$. Consider iteration $j$ of the DomToIP algorithm. We have $z\leq x^{(\ell+1)}\leq x^{(j)}$.
%\cindy{I certainly believe you and Bob on correctness, and I was convinced of the correctness of (and wrote a bunch of the proofs for) the earlier version.  But while I have a little time, I'll point out places where I was confused (and hence a reader might be). This last statement I don't see right away.  It doesn't seem that $x^{(\ell+1)}\leq x^{(j)}$ in general when $j < \ell + 1$. It seems that in general, for a fractional element $x_q$, DomToIP can round the $q$th element either up or down depending on whether it's OK to round down (future feasibility assured).  So it seems that $x^j$, which is more fractional, could have some elements smaller and some elements larger than the more integral $x^{(\ell+1)}$.  This confusion carries forward to the assertion that $z$ is a feasible solution to $\HelperLP(x^{(j-1)})$.  Suppose that $x_q = .2$ for some $q < \ell$ and Dom2IP rounded the $q$th element up to $1$. $\HelperLP(x^{q})$ has a constraint $z_q \le 0.2$, but subsequent LPs, such as $\HelperLP(x^{(\ell)})$ have a constraint $x_q = 1$. So the later LP solutions may not be feasible for the earlier ones.}
We also have $x^{(j)}_j=1$ which implies when we solved $\HelperLP(x^{(j-1)})$ the optimal value was strictly larger than zero. However, $z$ is a feasible solution to $\HelperLP(x^{(j-1)})$ and gives an objective value of 0. This is a contradiction, so $z\in \IP^{(\ell+1)}$.

Now for the second case, assume $z^*_{\ell+1} > 0$. We set $x^{(\ell+1)}_{\ell+1}=1$. For each point $z\in \LP^{(\ell)}$ we have $z_{\ell+1} >0$, so for each $z\in \IP^{(\ell)}$ we have $z_{\ell+1}>0$, i.e. $z_{\ell+1}=1$. This means that $z\in \IP^{(\ell+1)}$, and $\IP^{(\ell+1)} \neq \emptyset$.
